\chapter{Introduction}\label{ch:intro}

Capacity planning is concerned with predicting how system performance changes as workloads grow and resources become constrained. In modern microservice-based systems, this task is increasingly challenging: service times shift under CPU pressure, tail latencies become heavy-tailed, and saturation points move as demand fluctuates. Understanding these effects is essential for anticipating overload, ensuring responsiveness, and making informed resource-allocation decisions.

Although courses such as COMP9334 introduce the foundations of queueing theory, operational laws, and discrete-event simulation (DES), students rarely encounter a setting where these behaviours can be observed directly. Without empirical measurements, key ideas—such as utilisation growth, queue formation, tail-latency amplification, and the onset of saturation—remain abstract. This disconnect makes it difficult to appreciate how analytical models behave when applied to real systems exhibiting variability, contention, and non-linear delay characteristics.

This thesis addresses that gap by constructing a controlled experimental platform around a simple microservice exposed to varying workloads and CPU constraints. The resulting performance traces—service times, latency distributions, utilization, and indicators of queue buildup—form the empirical basis for the modeling work. Using these traces, the thesis develops a hybrid modeling approach that combines machine-learning methods with DES: ML learns how service-time distributions evolve under resource pressure, while DES uses these learned distributions to simulate queueing dynamics, tail behaviour, and saturation thresholds. This combination captures phenomena that neither classical queueing models nor purely data-driven methods can fully explain in isolation.

By comparing empirical measurements with analytical predictions, DES outputs, and learned performance models, the study shows how classical theory and modern data-driven techniques complement each other in contemporary capacity planning. The unified case study provides students with a realistic environment in which they can observe performance behaviour first-hand, validate theoretical concepts, and explore predictive approaches to scaling, resource allocation, and overload prevention. We also address the fact that existing work on container benchmarking provides measurements but not predictive modeling; DES research provides modeling frameworks but lacks integration with real containerized systems; and ML-assisted DES exists but not in the context of computing performance or microservices. This thesis develops an empirical–ML–DES hybrid capable of forecasting performance degradation and tail behaviour under changing workload and resource conditions—a capability neither traditional queueing models nor current ML systems provide alone. Although designed to run on student machines, the framework addresses a genuinely open industrial problem: predicting how microservice performance evolves under CPU contention and fluctuating demand.

The remainder of this thesis unfolds in a structured progression. First, the background chapter introduces the foundational concepts needed to interpret modern capacity-planning behaviour—covering utilisation laws, queueing models, operational analysis, and simulation principles. Next, a focused literature review surveys industrial practises and prior academic work, highlighting how contemporary systems are monitored, modelled, and scaled, and identifying the gaps that motivate the framework developed here. This is followed by the detailed project plan and methodology, which outline how empirical traces were collected, how modeling components were designed, and how the hybrid DES–ML approach will be implemented.